---
output:
  html_document:
    df_print: paged
editor_options: 
  chunk_output_type: console
---

# BIOEE 3550/6550 : **Data analysis and visualization in ecology and environmental science**

### ***Spring 2025***

### Instructor: **Xiangtao Xu** ( ✉️ [xx286\@cornell.edu](mailto:xx286@cornell.edu){.email})

### Teaching Assistant: **Yixin Ma** (✉️ [ym524\@cornell.edu](mailto:ym524@cornell.edu){.email})

## [Homework 0]{style="color:royalblue"} *Hello World*

### 0. Configure R, RStudio, Rmarkdown

[*(Please complete before the class)*]{style="color:red"}

For R users, we are going to use Rstudio and Rmarkdown as the major platform for most of the numerical labs and homework. R markdown is a simple and easy to use plain text language used to combine your R code, results from your data analysis (including plots and tables) and written commentary into a single nicely formatted and reproducible document.

#### 0.1 Install R/Rstudio

First, install R and Rstudio (if you have not done that yet) following [this link](https://posit.co/download/rstudio-desktop/).

#### 0.2 Install Rmarkdown and other necessary packages

Then, run the following command in your Rstudio

```{r install-rmarkdown, eval=FALSE, include=TRUE}
# Install from CRAN
install.packages('rmarkdown', dep = TRUE)
library(rmarkdown)
install.packages('microbenchmark', dep = TRUE)
library(microbenchmark)
```

#### 0.3 Basics of Rmarkdown

Rmakrdown allows for inserting code blocks within markdown texts using \`\`\`. Please Go over Chapters `8.3`, `8.4`, and `8.5` of the [intro2r](https://intro2r.com/get-started-with-r-markdown.html) documentation

Try executing this chunk by clicking the *Run* button within the chunk or by placing your cursor inside it and pressing *Ctrl+Shift+Enter*.

```{r}
print("Hello World")
```

### 1. Print the length of rainfall data

We have provided codes to read an array of daily rainfall data below. Please run the codes and print the length of the data array (i.e. how many data points there are) based on examples during the lecture (or with the help of online or AI resources).

```{r}
# load the rainfall data
rainfall <- scan("./HM0_rainfall.txt")
```

```{r}

length(rainfall) |> 
  print()

```

### 2. Find the number of rainy days

Please reproduce the example presented in the first lecture on finding the number of rainy days (the first one with loop).

```{r}
# creates a counter object to be overwritten with the number of rainy days
rainy_days <- 0

# loop counts days
for (i in 1:length(rainfall)) {
  i_rain <- rainfall[i]
  
  # 'if' function adds to the counter when there is a value above 0 (i.e., there is rain)
  if (i_rain > 0) {
    rainy_days <- rainy_days + 1
  }
}

print(rainy_days)

```

### 3. Find the longest dry period (consecutive dry days)

Please reproduce the example presented in the first lecture.

```{r}

find_dry_period <- function(rainfall_input, start_index) {
  if (rainfall_input[start_index] > 0) {
    
    # If the input day is rainy (>0), then the longest dry period has to be 0
    dry_period_length <- 0
    
  } else {
    
    # If the input day is dry (0), then the loop will go over the data until the next wet day
    array_length <- length(rainfall_input)
    
    for (i in start_index:array_length) {
      
      # Breaks out of the loop if we encounter a wet day
      if (rainfall_input[i] > 0) {
        break
      }
    }
    
    # Then: count the number of days in the dry period
    
    if (i == array_length & rainfall_input[i] == 0) {
      
      # If the last day in a dry period is also the last record in the dataset
      # then the dry period length is the # of days between the index and the last record
      dry_period_length = i - start_index + 1
      
    } else {
      
      dry_period_length = i - start_index
      }
  }
  
  #finish the function
  return(dry_period_length)
}


# Find the length of consecutive dry days for each day, then store them in a list
# Make a list full of 0s that will get overwritten in the for loop below
dry_period_list <- numeric(length(rainfall))

# Fill the above list with values using the function written above
for (i in 1:length(rainfall)) {
  dry_period_list[i] <- find_dry_period(rainfall, i)
}


# Print the list
print(dry_period_list)

# Find the maximum value (longest dry period)
max(dry_period_list) |> 
  print()

```

### 4. Compare different vectorized and un-vectorized calculations

Please reproduce the example presented in the first lecture.

```{r}
# Unvectorized version
get_rainy_days <- function(rainfall_input) {
  rainy_days <- 0
  
  # Loop over each index of the rainfall array
  for (i in 1:length(rainfall)) {
    # Access the rainfall value for the current index
    i_rain <- rainfall[i]
    
    # Check if the current day has rain
    if (i_rain > 0) {
      # Increment the number of rainy days
      rainy_days <- rainy_days + 1
    }
  }
  
  return(rainy_days)
}


# Vectorized version
get_rainy_days_vectorised <- function(rainfall_input) {
  rainy_days <- sum(rainfall_input > 0)
  return(rainy_days)
}


# Check whether they give the same results
rainfall |> 
  get_rainy_days() |> 
  print()

rainfall |> 
  get_rainy_days_vectorised() |> 
  print()


```

```{r}
# Time the unvectorized and vectorized versions
timing_results <- microbenchmark(
  unvectorised = get_rainy_days(rainfall),
  vectorised = get_rainy_days_vectorised(rainfall),
  times = 1000
)


# Print the timing results
timing_results |> 
  print()

```
### 5. Find the longest wet period (consecutive wet days)

Based on the examples in the lecture, try to find the longest wet period in the data set. Please include appropriate annotations/comments for your codes

```{r}
# Find the longest wet period (consecutive wet days)

```